---
title: "501_graph_redo"
author: "Penny Kahn"
date: "9/26/2019"
output: html_document
---

```{r}
library(tidyverse)
library(ggplot2)
library(wesanderson)
library(ggthemes)
library(grid)
```

```{r}
pat_bar<-read.csv("501_graph_attempt1.csv")
```
```{r}
ggplot(pat_bar, aes(x=alga, y=no_genes, fill=species, color=affiliation))+
  geom_col(stat="identity", position=position_dodge())+
  theme_bw()
```

```{r}
pat_bar2 <- pat_bar %>%
  mutate(sp_affiliation = str_c(species, affiliation, sep = ", ")) %>%
  group_by(alga, species) %>% 
  mutate(prop = no_genes/sum(no_genes))
```

```{r}
ggplot(pat_bar2, aes(x=alga, y=no_genes, fill=species, alpha=affiliation))+
  geom_col(stat="identity", position=position_dodge())+
  scale_fill_manual(values=c("cornflowerblue", "gold"))+
  scale_alpha_discrete(range=c(0.4,1.0))+
  theme_minimal()
```
maybe figure out how to put these next to each other but keep them stacked

...wait thats exactly what the other paper did you idiot
```{r}
ggplot(pat_bar2, aes(x=alga, y=no_genes, fill=species, alpha=affiliation))+
  geom_col(stat="identity", position=position_fill())+
  scale_fill_manual(values=c("cornflowerblue", "gold"))+
  scale_alpha_discrete(range=c(0.4,1.0))+
  theme_minimal()
```

```{r}
pie<-read.csv("501_graph_attempt2.csv")
```

```{r}
pie %>% 
  select(species, location, number) %>% 
  group_by(species) %>% 
  mutate(prop = number/sum(number)) %>% 
  ggplot(aes(x=reorder(location, -prop), y=prop, fill=species))+
    geom_col(stat="identity", position=position_dodge())
```
```{r}
pie %>% 
  select(species, location, number) %>% 
  group_by(species) %>% 
  mutate(prop = number/sum(number)) %>% 
  ggplot(aes(x=species, y=prop, fill=reorder(location, -prop)))+
    geom_col(stat="identity", position=position_dodge())+
    geom_text(aes(label=number), vjust=-.5, position = position_dodge(0.9), size=3.5)+
    scale_fill_brewer(palette = "Accent")+
    theme_minimal()+
    theme(legend.position = c(0.85, 0.75))
```
OMG wesanderson palettes:

5 colors
#>  [1] "BottleRocket1"  "BottleRocket2"  "Rushmore1"      "Rushmore"      
#>  [5] "Royal1"         "Royal2"         "Zissou1"        "Darjeeling1"   
#>  [9] "Darjeeling2"    "Chevalier1"     "FantasticFox1"  "Moonrise1"     
#> [13] "Moonrise2"      "Moonrise3"      "Cavalcanti1"    "GrandBudapest1"
#> [17] "GrandBudapest2" "IsleofDogs1"    "IsleofDogs2"

```{r}
library(wesanderson)
pie %>% 
  select(species, location, number) %>% 
  group_by(species) %>% 
  mutate(prop = number/sum(number)) %>% 
  ggplot(aes(x=species, y=prop, fill=reorder(location, -prop)))+
    geom_col(stat="identity", position=position_dodge())+
    geom_text(aes(label=number), vjust=-.5, position = position_dodge(0.9), size=3.5)+
    scale_fill_manual(values=wes_palette(n=5, name="Moonrise3"))+
    theme_minimal()+
    theme(legend.position = c(0.85, 0.75))
```
```{r}
library(wesanderson)
library(ggthemes)
pie %>% 
  select(species, location, number) %>% 
  group_by(species) %>% 
  mutate(prop = number/sum(number)) %>% 
  ggplot(aes(x=species, y=prop, fill=reorder(location, -prop)))+
    geom_col(stat="identity", position=position_dodge())+
    geom_text(aes(label=number), vjust=-.5, position = position_dodge(0.9), size=3.5)+
    labs(x="Species", y="Proportion")+
    scale_fill_manual(values=wes_palette(n=5, name="Royal2"))+
    geom_rangeframe() +
    theme_tufte()+
    theme(legend.position = c(0.85, 0.75), legend.title = element_blank())
```
```{r}
pie %>%
  select(species, location, number) %>%
  group_by(species) %>%
  mutate(prop = number/sum(number)) %>%
  ggplot(aes(x=species, y=prop, fill=reorder(location, -prop)))+
    geom_col(stat="identity", position=position_dodge())+
    geom_text(aes(label=number), vjust=-.5, position = position_dodge(0.9), size=3.5)+
    labs(x="", y="Proportion")+
    scale_fill_manual(values=wes_palette(n=5, name="Moonrise3"))+
    geom_rangeframe()+
    theme_tufte()+
    theme(legend.position = c(0.85, 0.75), legend.title = element_blank(), axis.text.x=element_text(size=14, face="italic", color="black"), axis.title = element_text(size=12))+
    annotate("text", x = 1.1, y = 0.35,
              label = "Total alga-derived proteins:\n353",
              family = "serif")+
    annotate("text", x = 2.1, y = 0.35,
              label = "Total alga-derived proteins:\n508",
              family = "serif")
```

```{r}
t353 <- textGrob("Total alga-derived proteins:\n353", gp=gpar(fontsize=9, fontfamily="serif"))
t508 <- textGrob("Total alga-derived proteins:\n508", gp=gpar(fontsize=9, fontfamily="serif"))

pie %>%
  select(species, location, number) %>%
  group_by(species) %>%
  mutate(prop = number/sum(number)) %>%
  ggplot(aes(x=species, y=prop, fill=reorder(location, -prop)))+
    geom_col(stat="identity", position=position_dodge())+
    geom_text(aes(label=number), vjust=-.5, position = position_dodge(0.9), size=3.5)+
    labs(x="", y="Proportion")+
    scale_fill_manual(values=wes_palette(n=5, name="Moonrise3"))+
    geom_rangeframe() +
    theme_tufte()+
    theme(legend.position = c(0.85, 0.75), legend.title = element_blank(), axis.text.x=element_text(size=14, face="italic", color="black"), axis.title = element_text(size=12))+
    annotation_custom(t353, xmin=0.5, xmax=1.5, ymin= -0.13, ymax=-0.11)+
    annotation_custom(t508, xmin=1.5, xmax=2.5, ymin= -0.13, ymax=-0.11)+
    coord_cartesian(clip = "off")
```
